Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-16 text/manuscript-full-KS.qmd
worked through front end

@RN5007[p. 402] define quantification as "the production and communication of numbers." The effects of quantification are described by the three themes of reactivity (people and orginizations change behavior in response to salient quantitative measures), discipline (quantification disciplines actors to react in particular ways), and authority (quantification often weakens discretion of local decision-makers). Sociological studies on quantification have predominantly explored how numbers are produced, used, and interpreted across institutional contexts such as administration, democratic rule, economics, personal life and education, with a particular emphasis on rankings [@RN5008; @RN5003; @RN5005]. In the field of critical data studies, quantification is central to scholarship that explores how computation and algorithms structure social opportunities [e.g., @RN4772; @RN4849]. Our study is grounded in the systematic investigation of "interactions between different quantification regimes" across sociology and critical data studies, as called for by @RN5008[p. 239], to explore the role of third-party servicers and products in the sorting of students into colleges. In particular, the discussions of correlation and homophily by @RN4975 show how social reproduction is encoded in algorithms that direct future resources. We explore how correlation and homophily were incorporated into the construction and application of Geomarkets within the Market Segment Model [@RN4982].

@@

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homophily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "social networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@RN4975, p. 94]. We observe similar phenomena in our case. Based on a network analysis of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

@@

In 1978, set against the backdrop of impending college-age demographic decline, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. The project goal became the creation of the Market Segment Model, which had the lofty ambition of predicting student demand for any college from any locality. However, "to gain a truly comprehensive view of the collegiate enrollment market, we needed a database that described most institutions and most students" [@RN4982, p. x]. The group approached College Board. "Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential, efforts which had faltered largely because the smallest deomgraphic unit used in these analyses was the state" [@RN4982, p. x].

Although the analyses of @RN4982 are based on the score-sending behavior of SAT test-takers, the authors had great respect for admissions officers and conceived of the project as an effort to "capture and quantify" (p. 11) their knowledge:

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting (p.9)\ldots We have begun with this celebration of storytelling\ldots \[because\] we believe that the intuitions of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research is based upon listening carefully to what admissions officers have to say (pp. 9-10).

The "initial task" was to define geographic boundaries "in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982,p. 4]. Quoting an admissions officer, @RN4982 \[p. 11\] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, @RN4982 [p. 11] created "three types of boundaries -- region, state, and community. The initial regions in the project were New England, Middle States, and the South. Next, they "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982 [pp. 11-12] described the creation of Geomarket borders only briefly:

@@

Having defined geography, the Market Segment Model sought to predict how student demand for a particular college varies by Geomarket, based on the characteristics of households. The model created a straightforward but consistent classification system that grouped SAT test-takers into one of four different *market segments* -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc.

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.^[For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.^[Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

**Correlations**. @RN4982 [Chapter 3], *A Sense of Place: Students, Families, and Communities*, identifies the student characteristics correlated with being in the local, in-state, regional, or national market segment. The analyses identify four variables -- educational aspirations, parental education, scholastic aptitude, and family income -- that predict score sending behavior, both individually and in combination. Because these four variables "reflect the basic social patterns of the nation, it would have been surprising if these were not the four social variables that best explained the patterns of college choice" [@RN4982, p. 33].<!--
and "The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice" (pp. 34-35). \[QUOTES REDUNDANT\]
-->

The thesis of Chapter 3 is that student demand for higher education is a function of social origin. For the authors (p. 33) state that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and that "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982 \[p. 33\] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time."\[\^7\] @RN4982 [p. 42] conclude that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations."

The authors also find that these four variables predict student score-sending behavior at the Geomarket-level. However, Geomarkets differ in the relative abundance of students with particular socioeconomic characteristics, which has practical implications for recruiting. Therefore, @RN4982 [p. 44] recommend that colleges target Geomarkets with desirable compositions of socioeconomic characteristics in order to reach students from desired student market segments:

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one another.^[This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."] Based on these analyses, @RN4982 \[p. 46\] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

@@

The discussion of competition by @RN4982 exemplifies the concerns about correlation and homophily described by @RN4975. A correlational analysis of 1980 SAT score-sending patterns finds competition between colleges is defined by socioeconomic homophily. This homophily is presented as a naturally occurring phenomenon. Given these findings, @RN4982 recommend that colleges should target Geomarkets that contain a critical mass of students interested in peer-colleges, information that can be discerned from the Institutional Profiles (Appendix @tbl-sample-institutional-profile). In itself, @RN4982's Market Segment Model is a social science depiction of student demand -- akin to the status attainment model -- that does not consider historic, structural inequalities that cause observed patterns. However, by inscribing the Market Segment Model and Geomarkets into the EPS software, College Board amplified structural inequalities that contributed to homophily observed in 1980 SAT score-sending beahivor. In this way, commodification of the Market Segment Model and Geomarkets "engineers homophily" [@RN4975].

@@

Quantification disciplines actors to react in particular ways. EPS software may "discipline" [@RN5007] colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.^[Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

The effects of quantification also lead to changes in decision-making authority, frequently reducing the discretion of local decision-makers. @RN4982 sought to develop a concrete, data-driven framework -- The Market Segment Model -- that replicates the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools.

Finally, we suspect that Geomarkets affect how admissions offices organize the recruiting territories assigned to each admissions officer. Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket. In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago \[ADD TO APPENDIX\], of Lake Forest College, of Saint Mary's College \[ADD TO APPENDIX CRYSTAL\] <!-- uc san diego appears to use geomarkets to organize high school visits: inspect the html in this page https://beatriton.ucsd.edu/register/requestschoolvisit -->

Since @RN4982 developed The Market Segment Model, Geomarkets have been central to EPS software and play a broader role in shaping how many colleges organize recruiting territories. We therefore frame our investigation of Geomarkets as a case study of quantification to explore the extent to which this geographical building block is associated with race and class. We first address (RQ1) what is the socioeconomic and racial variation between Geomarkets and how does this variation change over time. <!--# Although we cannot gain access to EPS software, by focusing on metropolitan areas that are associated with three or more Geomarkets-->

@RN4982 viewed demand for higher education as a function of class and developed Geomarkets with an eye towards identifying geographic areas that differed from one another in terms of class composition. Given the extent of class-based residential segregation in the U.S., we therefore expect substantial socioeconomic inequality between Geomarkets in large metropolitan areas.

@@

We also expect substantial racial inequality between Geomarkets within metropolitan areas. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race. U.S. cities are characterized by extreme historic and contemporary residential racial segregation [@RN4779]. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, although geomarket borders may have been drawn along class divides, a strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the countours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets. <!--#  Structures built upon racialized structures are racialized structures [RN4786].-->

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.^[For public universities that purchased lists, 80% purchase more than 50,000 names annually. @RN4896 reports that student lists were the top expenditure item in the undergraduate recruiting budget for both private and public institutions in 2022, with the average public institution allocating 15% of its budget to purchasing names.] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

Student lists are positively associated with student outcomes. A College Board research report by @RN4739 compared SAT test-takers who opted into the College Board Student Search Service -- allowing colleges to purchase their contact information -- to those who opted out. After controlling for covariates (e.g., SAT score, parental education, school fixed effects), 41.1% of students who participated in Search attended a 4-year college compared to 32.8% of students who opted out, an 8.3 percentage point difference and a 25.3 percent change -- $(41.1–32.8)/32.8$ -- in the relative probability.^[Participating in Search was associated with a larger change in the relative probability of attending a 4-year college for Black students (24.5%) and Hispanic students (34.4%) than White students (21.6%), and a larger change for students whose parents did not attend college (40.6%) than those whose parents had a bachelor’s degree (18.9%). Leveraging a natural experiment in College Board student list purchases, @RN4894 find that purchasing a prospect profile increases the probability that the student will apply to and enroll at the purchasing college, with larger effects for Black, Hispanic, and lowincome students.] @RN4739 also found that participating in Search was strongly associated with obtaining a BA in four years, particularly for sub-groups that were historically excluded from college.^[20.6% of students who participated in Search obtained a BA in four years compared to 15.7% of students who opted out, representing a 31.2% (=(20.6-15.7)/15.7) increase in the relative probability of graduation. Furthermore, the relative increase in the probability of obtaining a BA was higher for Black (40.3%), Hispanic (43.3%), and Native American/Alaska Native students (48.3%) than it was for White (27.9%) and Asian (28.2%) students. The relative increase was also higher for students whose parents did not attend college (50.0%) than it was for students whose parents had a BA (25.3%).]

@@

> Imagine for the moment that you recruit for a college that draws most of its students from regional or national segments. Where would you concentrate your energies? Ideally, you would seek communities \[Geomarkets\] with a high proportion of students already predisposed toward institutions such as your own. The Market Segment Model would provide this information through segment percentages for the community \[Geomarket\] in question. Further classification of students by social attributes allows you to identify a group for mailings or recruiting\ldots If you were to recruit in Boston, only about two out of every ten students with fewer than two attributes would likely listen, while slightly less than half of the students with two or more attributes would be receptive\ldots. Your efforts would surely be better directed toward \ldots Manchester, Hartford, and Fairfield County \ldots Indeed, in Fairfield County alone you could reach more than 40 percent of your "primary target" population -- that is, students with a greater than 75-percent probability of concentrating their college choices among institutions like your own.

>>>>

@RN5007[p. 402] define quantification as "the production and communication of numbers." <!-- The effects of quantification are described by the three themes of reactivity (people and orginizations change behavior in response to salient quantitative measures), discipline (quantification disciplines actors to react in particular ways), and authority (quantification often weakens discretion of local decision-makers).--> The effects of quantification are described by the concepts reactivity, discipline, and authority [@RN5007]. Sociological studies on quantification have predominantly explored how numbers are produced, used, and interpreted across institutional contexts such as administration, democratic rule, economics, personal life and education, with a particular emphasis on rankings [@RN5008; @RN5003; @RN5005]. In the field of critical data studies, quantification is central to scholarship that explores how computation and algorithms structure social opportunities [e.g., @RN4772; @RN4849]. Our study is grounded in the systematic investigation of "interactions between different quantification regimes" across sociology and critical data studies, as called for by @RN5008[p. 239], to explore the role of third-party servicers and products in the sorting of students into colleges. In particular, the discussions of correlation and homophily by @RN4975 show how social reproduction is encoded in algorithms that direct future resources. We explore how correlation and homophily were incorporated into the construction and application of Geomarkets within the Market Segment Model [@RN4982].

@@

@RN4975 problematizes the idea that homophily is a naturally occurring phenomenon. Because network science models -- like correlational models above -- typically do not observe how historical structures affect relationships, they "erase historical contingencies, institutional discrimination, and economic realities" [@RN4975, p. 95] that cause behavior consistent with homophily. Second, in commercial social networks, homophily is more than an assumption; rather, it is programmed into algorithms that create connections between users. Thus, "networks create and spawn the reality they imagine; they become self-fulfilling prophecies" [@Chun2018Queerying, p. 66]. We observe similar phenomena in our case. Based on a network analysis of 1980 SAT score-sending, @RN4982 concludes that like-colleges compete for like-students (i.e., students of similar social origin). In turn, @RN4982 reasons that colleges should target the Geomarkets and high schools targeted by peer colleges. This logic is subsequently programmed into EPS software that recommends which Geomarkets and high schools to recruit from. Thus, homophily observed on the demand side -- which is a function of historic structural inequality –- is programmed into the supply-side.

@@

In 1978, set against the backdrop of impending college-age demographic decline, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. To answer these questions, Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. The project goal became the creation of the Market Segment Model, which had the lofty ambition of predicting student demand for any college from any locality. However, "to gain a truly comprehensive view of the collegiate enrollment market, we needed a database that described most institutions and most students" [@RN4982, p. x]. The group approached College Board. "Coincidentally, the Board was reviewing its own efforts to help colleges estimate their enrollment potential" [@RN4982, p. x]. Although the analyses of @RN4982 are based on the score-sending behavior of SAT test-takers, the authors had great respect for admissions officers and conceived of the project as an effort to "capture and quantify" (p. 11) their knowledge:

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting\ldots We have begun with this celebration of storytelling\ldots \[because\] we believe that the intuitions of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research is based upon listening carefully to what admissions officers have to say (pp. 9-10).

The "initial task" was to define geographic boundaries "in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982, p. 4]. Quoting an admissions officer, @RN4982[p. 11] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, @RN4982[p. 11] created "three types of boundaries -- region, state, and community." The initial regions in the project were New England, Middle States, and the South. Next, they "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982 [pp. 11-12] described the creation of Geomarket borders only briefly:

@@

Having defined geography, the Market Segment Model sought to predict how student demand for a particular college varies by Geomarket, based on the characteristics of households. The model created a classification system that grouped SAT test-takers into one of four different *market segments* -- local, in-state, regional, and national -- based on SAT score-sending behavior. For a given student, each college that receives a score from the student can be defined as "local" (college located in the the same Geomarket as the student), "in-state" (same state but different Geomarket), "regional" (same region but different state), or "national" (different region). In turn, a test-taker is categorized in the "local" market segment if they submit more SAT scores to local institutions than they do to in-state, regional, or national institutions. An "in-state" student submits more SAT scores to in-state institutions than they do to local, regional, or national institutions, etc. @RN4982[p. 4] comment that the Market Segment Model "was nothing more than a set of simple rules for disaggregating high school seniors into similar groups. The model worked because students, once so disaggregated, appeared to behave in remarkably consistent ways."

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail [CRYSTAL]. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.^[For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.^[Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.] Taken together, the Market Segment Profile shows a college which Geomarkets possess attractive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

**Correlations**. @RN4982 [Chapter 3], *A Sense of Place: Students, Families, and Communities*, investigates the student characteristics correlated with being in the local, in-state, regional, or national market segment. The analyses identify four variables -- educational aspirations, parental education, scholastic aptitude, and family income -- that predict score sending behavior, both individually and in combination. Because these four variables "reflect the basic social patterns of the nation, it would have been surprising if these were not the four social variables that best explained the patterns of college choice" [@RN4982, p. 33].<!--
and "The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice" (pp. 34-35). \[QUOTES REDUNDANT\]
-->

The thesis of this chapter is that student demand for higher education is a function of social origin. For example, the authors state (p. 33) that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and that "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982[p. 33] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time." They conclude (p. 42) that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations."

The authors also find that these four variables predict student score-sending behavior at the Geomarket-level. However, Geomarkets differ in the relative abundance of students with particular socioeconomic characteristics, which has practical implications for recruiting. Therefore, @RN4982[p. 44] recommend that colleges target Geomarkets with desirable compositions of socioeconomic characteristics in order to reach students from desired student market segments:

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one another.^[This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."] Based on these analyses, @RN4982[p. 46] state that competition between colleges is characterized by homophily: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and the socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe observed patterns as a natural process in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

@@

The discussion of competition by @RN4982 exemplifies the concerns about correlation and homophily described by @RN4975. A correlational analysis of 1980 SAT score-sending patterns finds competition between colleges is defined by socioeconomic homophily. This homophily is presented as a naturally occurring phenomenon. Given these findings, @RN4982 recommend that colleges should target Geomarkets that contain a critical mass of students interested in peer colleges, information that can be discerned from the Institutional Profiles (Appendix @tbl-sample-institutional-profile). In itself, @RN4982's Market Segment Model is a social science depiction of student demand -- akin to the status attainment model -- that does not consider historic, structural inequalities that cause observed patterns. However, by inscribing the Market Segment Model and Geomarkets into the EPS software, College Board amplified structural inequalities that contributed to homophily observed in 1980 SAT score-sending beahivor. Drawing from @RN4975, commodification of the Market Segment Model and Geomarkets engineers homophily.

@@

Quantification "disciplines" [@RN5007] actors to react in particular ways. EPS software may discipline colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.^[Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges. Anecdotally, the salience of Geomarkets and EPS software affected the decision-making authority of admissions officers and the territories that admissions officers were assigned to.^[Quantification often reduces the decision-making authority of local actors [@RN5007]. The Market Segment Model sought to replicate the aggregate knowledge of local admissions officers. Once this knowledge was quantified and commodified onto a CD-ROM, the local expertise of admissions officers becomes less valuable. The EPS product increases the ability of a college admissions leader -- working with College Board staff or an EM consultant -- to plan recruiting efforts centrally. In background conversations, EM professionals told us that EPS software enabled colleges to plan travel without relying on admissions officers having strong local knowledge of their territories. However, EPS software often recommended visiting the same sets of affluent, high-achieving high schools that were receiving visits from other colleges. We were told that savvier, quantitatively-adept admissions offices used EPS to visit schools that EPS recommended ignoring because there would be less competition for the good students who attended these high schools. Finally, Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it remains common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket. In states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago [ADD TO APPENDIX], of Lake Forest College, of Saint Mary's College [ADD TO APPENDIX CRYSTAL]] <!-- uc san diego appears to use geomarkets to organize high school visits: inspect the html in this page https://beatriton.ucsd.edu/register/requestschoolvisit -->

Because Geomarkets are foundational to EPS software, our first research question is interested in the extent to which this geographical building block is associated with race and class. We ask (RQ1), what is the socioeconomic and racial variation between Geomarkets and how does this variation change over time? @RN4982 viewed demand for higher education as a function of class. They developed 142 Geomarkets borders with an eye towards identifying geographic areas that differed from one another in terms of class composition, while [@RN5016] created the remaining Geomarket borders with an eye towards distinguishing geographies based on college-going behaviors. Considering the extent of class-based residential segregation in the U.S. [@reardon2011], we expect substantial socioeconomic inequality between Geomarkets in large metropolitan areas.

@@

We also expect substantial racial inequality between Geomarkets within metropolitan areas. Interestingly, the Market Segment Model is explicitly based on socioeconomic stratification, but @RN4982 do not mention race. U.S. cities are characterized by extreme historic and contemporary residential racial segregation [@RN4779]. Unless designers intentionally consider racial segregation, selection devices that categorize people based on geographic location are likely to reproduce historical race-based inequality in opportunity [@RN4975]. Second, although geomarket borders may have been drawn along class divides, a strong correlation exists between race and wealth [@RN4884]. Third, Geomarket borders may have been drawn in a way that follows the contours of racial segregation in residential housing. Examples include: the "South and South Central Los Angeles" geomarket (CA21); the "City of Oakland" geomarket (CA07), which is surrounded by the "Alameda County excluding Oakland" geomarket (CA08); and the "Wayne County Detroit" geomarket (MI01), which is surrounded by the "Detroit's Northern Suburbs" (MI02) and "Ann Arbor" (MI03) geomarkets. <!--#  Structures built upon racialized structures are racialized structures [RN4786].-->

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.^[For public universities that purchased lists, 80% purchase more than 50,000 names annually. @RN4896 reports that student lists were the top expenditure item in the undergraduate recruiting budget for both private and public institutions in 2022, with the average public institution allocating 15% of its budget to names.] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

Student lists are positively associated with student outcomes. A College Board research report by @RN4739 compared SAT test-takers who opted into the College Board Student Search Service -- allowing colleges to purchase their contact information -- to those who opted out. After controlling for covariates (e.g., SAT score, parental education, school fixed effects), 41.1% of students who participated in Search attended a 4-year college compared to 32.8% of students who opted out, an 8.3 percentage point difference and a 25.3 percent change -- $(41.1–32.8)/32.8$ -- in the relative probability.^[Leveraging a natural experiment in College Board student list purchases, @RN4894 find that purchasing a prospect profile increases the probability that the student will apply to and enroll at the purchasing college, with larger effects for Black, Hispanic, and low-income students.] <!--^[Participating in Search was associated with a larger change in the relative probability of attending a 4-year college for Black students (24.5%) and Hispanic students (34.4%) than White students (21.6%), and a larger change for students whose parents did not attend college (40.6%) than those whose parents had a bachelor’s degree (18.9%).] --> @RN4739 also found that participating in Search was strongly associated with obtaining a BA in four years.^[20.6% of students who participated in Search obtained a BA in four years compared to 15.7% of students who opted out, representing a 31.2% (=(20.6-15.7)/15.7) increase in the relative probability of graduation.] <!-- , particularly for sub-groups that were historically excluded from college.^[. Furthermore, the relative increase in the probability of obtaining a BA was higher for Black (40.3%), Hispanic (43.3%), and Native American/Alaska Native students (48.3%) than it was for White (27.9%) and Asian (28.2%) students. The relative increase was also higher for students whose parents did not attend college (50.0%) than it was for students whose parents had a BA (25.3%).]-->

@@

> Imagine that you recruit for a college that draws most of its students from regional or national segments. Where would you concentrate your energies? Ideally, you would seek communities \[Geomarkets\] with a high proportion of students already predisposed toward institutions such as your own. The Market Segment Model would provide this information through segment percentages for the community in question. Further classification of students by social attributes allows you to identify a group for mailings or recruiting\ldots If you were to recruit in Boston, only about two out of every ten students with fewer than two attributes would likely listen \ldots Your efforts would surely be better directed toward \ldots Manchester, Hartford, and Fairfield County \ldots Indeed, in Fairfield County alone you could reach more than 40 percent of your "primary target" population. <!-- -- that is, students with a greater than 75-percent probability of concentrating their college choices among institutions like your own. -->
